Secure Composition of Robust and Optimising Compilers

### 2023-06-22 popl/main.tex
Patch figure

\begin{center}
  \judgbox{\thelocmap(\src{\loc})=\loc}{,,Map an $\src{L_{\tmssafe}}$ location to a specification location $\loc$.''}
  \judgbox{\filterLtms{\src{\event}}=\event}{,,Project an $\src{L_{\tmssafe}}$ event $\src{\event}$ to a specification event $\event$.''}
  \typerule{$\src{L_{\tmssafe}}$-filter-context}{
    \src{\event[_b]}\not=\src{\lightning}
  }{
    \filterLtms{\src{\tup{\event[_b]}{\ctx}} = \emptyevent}
  }{ltms-filter-context}
  %
  \typerule{$\src{L_{\tmssafe}}$-filter-abort}{
  }{
    \filterLtms{\src{\lightning}} = \lightning
  }{ltms-filter-abort}
  %
%  \typerule{$\src{L_{\tmssafe}}$-filter-start}{
%  }{
%    \filterLtms{\src{\tup{Start}{\comp}}} = \emptyevent
%  }{ltms-filter-start}
  %
  \typerule{$\src{L_{\tmssafe}}$-filter-alloc}{
    \locmapsto{\src{\loc}}{\loc} &
    \src{n}=n
  }{
    \filterLtms{\src{\tup{Alloc\ \loc\ n}{\comp}}} = \tup{\ev{Alloc}\ \loc\ n}{\comp,\unlock}
  }{ltms-filter-alloc}
\end{center}
To prove that well-typed $\src{L_{\tmssafe}}$ programs are temporal memory safe, a map from $\src{L_{\tmssafe}}$ to specification events (see \Cref{subsec:propdefs}) allows the use of \Cref{def:monsat}.
The map is mostly entirely straight-forward.
Most notably, it filters any action that the context does (\Cref{tr:ltms-filter-context}), since a context is free to do anything.
Of course, a context invoking $\src{\labort}$ to crash the program must not go undetected (\Cref{tr:ltms-filter-abort}).
As for the component, pre--events are essentially recolored to go from blue $\src{L_{\tmssafe}}$ events to the events of the most general trace model (\Cref{tr:ltms-filter-alloc}).
Events that carry locations make use of a location--map from blue $\src{L_{\tmssafe}}$ locations to the more abstract ones.

\begin{lemma}[$\src{L_{\tmssafe}}$-programs are \gls{tms}]\label{lem:wt:tmsmon}
  If $\progstepto{\src{\prog{\library_{\ctx}}{\library_{\comp}}}}{\src{\rtt{\statevar}{\valueexpr}}}{\src{\trace}}$, then $\montmssafe{\filterLtms{\src{\trace}}}$.\Coqed
\end{lemma}
\begin{theorem}[$\src{L_{\tmssafe}}$-programs are \gls{tms}]\label{thm:wt:tms}
  For any $\src{\library_{\comp}}$, $\rsat{\src{\library_{\comp}}}{\tmssafe}$ \Coqed
\end{theorem}

@@

The $\Ltrg$ language does not provide any guarantees with regards to memory safety.
The syntax of it extends the unified syntax for all languages (\Cref{subsec:cs:defs}) with pairs of natural numbers (\Cref{tr:e-proj1}) and two different kinds of dynamic checks.
One of those checks allows programmers to investigate wether a pointer is freed or not (\Cref{tr:e-ispoisoned-yes}).
The other check is a simple dynamic type check that evaluates to $\trg{0}$ if the check suceeds and to $\trg{1}$ otherwise (\Cref{tr:e-n-has-nat,tr:e-n-has-nattimesnat}).
The projection $\filterL{\trace}$ is entirely similar to $\filterLtms{\trace}$.
\MKin{add note about \gls{sms} projection}

@@

This paper chooses to introduce $\Lms$, despite it being exactly equal to $\Ltrg$.
The main purpose of this decision is to enhance readability.

\MKin{add note about \gls{sms} projection}

@@

\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
      \node (S) {$\src{L_{\tmssafe}}$};
      \node[right=1.0 of S] (T) {$\trg{L}$};
      \node[right=1.0 of T] (M) {$\irl{L_{\mssafe}}$};
      \node[below right=0.5 and 1.0 of M] (D0) {$\irl{L_{\mssafe}}$};
      \node[above right=0.5 and 1.0 of M] (C0) {$\irl{L_{\mssafe}}$};
      \node[right=1.0 of D0] (D1) {$\irl{L_{\mssafe}}$};
      \node[right=1.0 of C0] (C1) {$\irl{L_{\mssafe}}$};
      \node[right=4.5 of M] (O) {$\obj{L_{\scctsafe}}$};

      \draw[->] (S) to[sloped,above] node (tmsedge) {\text{\scriptsize\gls{tms}}} (T);
      \draw[->] (T) to[sloped,above] node {\text{\scriptsize\gls{sms}}} (M);
      \draw[->] (M) to[sloped,above] node {\text{\scriptsize\gls{dce}}} (D0);
      \draw[->] (M) to[sloped,above] node {\text{\scriptsize\gls{cf}}} (C0);
      \draw[->] (D0) to[sloped,above] node {\text{\scriptsize\gls{cf}}} (D1);
      \draw[->] (C0) to[sloped,above] node {\text{\scriptsize\gls{dce}}} (C1);
      \draw[->] (D1) to[sloped,above] node {\text{\scriptsize\gls{scct}}} (O);
      \draw[->] (C1) to[sloped,above] node {\text{\scriptsize\gls{scct}}} (O);

      % Sections
      \node[above=1.5 of tmsedge] (sectms) {{\scriptsize\Cref{subsec:cs:tms}}};
      \node[right=0.0 of sectms] (secsms) {{\scriptsize\Cref{subsec:cs:ms}}};
      \node[right=1.25 of secsms] (secopts) {{\scriptsize\Cref{subsec:cs:opts}}};
      \node[right=1.25 of secopts] (secscct) {{\scriptsize\Cref{subsec:cs:scct}}};

      \draw[dashed] ($(sectms)-(0.7,0)$) -- ($(sectms)-(0.7,1.75)$);
      \draw[dashed] ($(sectms)-(-0.7,0)$) -- ($(sectms)-(-0.7,1.75)$);
      \draw[dashed] ($(secsms)-(-0.7,0)$) -- ($(secsms)-(-0.7,1.75)$);
      \draw[dashed] ($(secscct)-(0.7,0)$) -- ($(secscct)-(0.7,0.75)$);
      \draw[dashed] ($(secscct)-(-0.7,0)$) -- ($(secscct)-(-0.7,1.25)$);
    \end{tikzpicture}
  \end{center}
  \caption{Visualisation of the optimizing compilation pipeline that attains a combination of \gls{ms} and \gls{cct}.}\label{fig:pipeline}
\end{figure}
This section defines several secure compilers, each of which robustly preserves a different property of interest (\Cref{fig:pipeline}).
The power of the framework (\Cref{sec:compprop,sec:compcomp}) is demonstrated by composing these compilers for a secure, but optimizing, compilation chain that robustly preserves \gls{ms} and \gls{cct}.
Concretely, the section starts with a base-line definition for the used languages (\Cref{subsec:cs:defs}) and extends that to define languages $\src{L_{TMS}}$ and $\trg{L}$.
For programmers using $\src{L_{TMS}}$, their mental model should be that their code is temporal memory safe and cryptographic constant time.
The former is proven to be enforced by the static semantics of $\src{L_{TMS}}$, where each well-typed program is temporal memory safe.
The latter is completely untyped.
Having languages defined, this section continues with a compiler $\cc{\src{L_{\tmssafe}}}{\trg{L}}$ and proves that $\cc{\src{L_{\tmssafe}}}{\trg{L}}$ robustly preserves \gls{tms} (\Cref{subsec:cs:tms}).
For readability purposes, it defines $\irl{L_{MS}}$ to be exactly the same as $\trg{L}$ and continues to present a compiler $\cc{\trg{L}}{\irl{L_{MS}}}$ that robustly preserves \gls{sms} (\Cref{subsec:cs:ms}).
Now leaveraging the power of the framework that this paper introduces (\Cref{sec:compcomp,sec:compprop}), a compiler $\rtp{\cc{\src{L_{TMS}}}{\irl{L_{MS}}}}{\mssafe}$ is attained.
The next steps in the compilation pipeline are two secure compilers $\cc[_{\gls{dce}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$ and $\cc[_{\gls{cf}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$ that perform \glsfirst{dce} and \glsfirst{cf}, respectively, to model an optimzing pipeline happening somewhere in the compilation chain (\Cref{subsec:cs:opts}).
Finally, this section defines the language $\obj{L_{\scctsafe}}$ that is, in contrast to the other languages, not cryptographic constant time by definition.
Concluding, a compiler $\cc{\irl{L_{MS}}}{\obj{L_{CCT}}}$ that robustly preserves \gls{cct} (\Cref{subsec:cs:scct}) is shown and composed with $\cc{\src{L_{TMS}}}{\irl{L_{MS}}}$ and the optimizing compilers $\cc[_{\gls{dce}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$ and $\cc[_{\gls{cf}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$.
In turn, this yields $\cc{\src{L_{TMS}}}{\obj{L_{CCT}}}$, a compiler that robustly preserves a combination of \gls{ms} and \gls{cct}.

@@

\MKin{backward sim}
\begin{theorem}[Compiler $\ccb$ is secure with respect to \gls{sms}]\label{thm:ccb:rtp:sms}
  $\rtp{\ccb}{\smssafe}$ % \Coqed
\end{theorem}

@@

\subsection{Optimizing Compilers}\label{subsec:cs:opts}
\MKin{rework this}

\MKin{compiler $\ccdce$ doing \gls{dce}}
\MKin{compiler $\cccf$ doing \gls{cf}}

\MKin{backward sims}
\MKin{rtp of ms}
\MKin{composition: rtp of ms, order independent}

@@

\MKin{compiler $\irl{L_{MS}}\to\obj{L_{\scctsafe}}$}
\MKin{x-lang relations}
\MKin{backtranslation}
\MKin{rtp of scct}

\MKin{composition: rtp of everything}

>>>>

\subsubsection{All Well-Typed $\Ltms$ Programs are Temporal Memory Safe}

\begin{center}
  \judgbox{\thelocmap(\src{\loc})=\loc}{,,Map an $\src{L_{\tmssafe}}$ location to a specification location $\loc$.''}
  \judgbox{\msfilterLtms{\src{\event}}=\event}{,,Project an $\src{L_{\tmssafe}}$ event $\src{\event}$ to a specification event $\event$.''}
  \typerule{$\src{L_{\tmssafe}}$-filter-context}{
    \src{\event[_b]}\not=\src{\lightning}
  }{
    \msfilterLtms{\src{\tup{\event[_b]}{\ctx}} = \emptyevent}
  }{ltms-filter-context}
  %
  \typerule{$\src{L_{\tmssafe}}$-filter-abort}{
  }{
    \msfilterLtms{\src{\lightning}} = \lightning
  }{ltms-filter-abort}
  %
%  \typerule{$\src{L_{\tmssafe}}$-filter-start}{
%  }{
%    \msfilterLtms{\src{\tup{Start}{\comp}}} = \emptyevent
%  }{ltms-filter-start}
  %
  \typerule{$\src{L_{\tmssafe}}$-filter-alloc}{
    \locmapsto{\src{\loc}}{\loc} &
    \src{n}=n
  }{
    \msfilterLtms{\src{\tup{Alloc\ \loc\ n}{\comp}}} = \tup{\ev{Alloc}\ \loc\ n}{\comp,\unlock}
  }{ltms-filter-alloc}
\end{center}
To prove that well-typed $\src{L_{\tmssafe}}$ programs are temporal memory safe, a map from $\src{L_{\tmssafe}}$ to specification events (see \Cref{subsec:propdefs}) allows the use of \Cref{def:monsat}.
The map is mostly entirely straight-forward.
Most notably, it filters any action that the context does (\Cref{tr:ltms-filter-context}), since a context is free to do anything.
Of course, a context invoking $\src{\labort}$ to crash the program must not go undetected (\Cref{tr:ltms-filter-abort}).
As for the component, pre--events are essentially recolored to go from blue $\src{L_{\tmssafe}}$ events to the events of the most general trace model (\Cref{tr:ltms-filter-alloc}).
Events that carry locations make use of a location--map from blue $\src{L_{\tmssafe}}$ locations to the more abstract ones.

\begin{lemma}[$\src{L_{\tmssafe}}$-programs are \gls{tms}]\label{lem:wt:tmsmon}
  If $\progstepto{\src{\prog{\library_{\ctx}}{\library_{\comp}}}}{\src{\rtt{\statevar}{\valueexpr}}}{\src{\trace}}$, then $\montmssafe{\msfilterLtms{\src{\trace}}}$.\Coqed
\end{lemma}
\begin{theorem}[$\src{L_{\tmssafe}}$-programs are \gls{tms}]\label{thm:wt:tms}
  For any $\src{\library_{\comp}}$, $\rsat{\src{\library_{\comp}}}{\tmssafe}$ \Coqed
\end{theorem}

@@

The $\Ltrg$ language does not provide any guarantees with regards to memory safety.
The syntax of it extends the unified syntax for all languages (\Cref{subsec:cs:defs}) with pairs of natural numbers (\Cref{tr:e-proj1}) and two different kinds of dynamic checks.
One of those checks allows programmers to investigate wether a pointer is freed or not (\Cref{tr:e-ispoisoned-yes}).
The other check is a simple dynamic type check that evaluates to $\trg{0}$ if the check suceeds and to $\trg{1}$ otherwise (\Cref{tr:e-n-has-nat,tr:e-n-has-nattimesnat}).
The projection $\msfilterL{\trace}$ is entirely similar to $\msfilterLtms{\trace}$.

@@

This paper chooses to introduce $\Lms$, despite it being exactly equal to $\Ltrg$.
The main purpose of this decision is to enhance readability.
One thing that is added to $\Lms$ is another projection to the most general trace model for reasoning about \gls{scct}.
Here, any information irrelevant to \gls{scct} is filtered.

\begin{center}
  \typerule{$\Lms$-filter-alloc}{
  }{
    \scctfilterLms{\irl{\tup{Get\ \loc\ n}{\comp}}} = \emptyevent
  }{lms-filter-get}
  %
  \typerule{$\Lms$-filter-abort}{
  }{
    \scctfilterLms{\irl{\lightning}} = \lightning
  }{lms-filter-abort}
\end{center}

@@

\begin{figure}[h]
  \begin{center}
    \begin{tikzpicture}
      \node (S) {$\src{L_{\tmssafe}}$};
      \node[right=1.5 of S] (T) {$\trg{L}$};
      \node[right=1.5 of T] (M) {$\irl{L_{\mssafe}}$};
      \node[below right=0.5 and 1.0 of M] (D0) {$\irl{L_{\mssafe}}$};
      \node[above right=0.5 and 1.0 of M] (C0) {$\irl{L_{\mssafe}}$};
      \node[right=3.0 of M] (E) {$\irl{L_{\mssafe}}$};
      \node[right=1.5 of E] (O) {$\obj{L_{\scctsafe}}$};

      \draw[->] (S) to[sloped] node[align=center,font=\scriptsize] (tmsedge) {\gls{tms}\\ \Cref{thm:cca:rtp:tms}} (T);
      \draw[->] (T) to[sloped] node[align=center,font=\scriptsize] {\gls{sms}\\ \Cref{thm:ccb:rtp:sms}} (M);
      \draw[->] (M) to[sloped] node[align=center,font=\scriptsize] {\gls{dce}\\ \Cref{thm:ccdce:rtp:ms}} (D0);
      \draw[->] (M) to[sloped] node[align=center,font=\scriptsize] {\gls{cf}\\ \Cref{thm:cccf:rtp:ms}} (C0);
      \draw[->] (D0) to[sloped] node[align=center,font=\scriptsize] {\gls{cf}\\ \Cref{thm:cccf:rtp:ms}} (E);
      \draw[->] (C0) to[sloped] node[align=center,font=\scriptsize] {\gls{dce}\\ \Cref{thm:ccdce:rtp:ms}} (E);
      \draw[->] (E) to[sloped] node[align=center,font=\scriptsize] {\gls{scct}\\ \Cref{thm:ccscct:rtp:scct}} (O);

      % Sections
      \node[above=1.0 of tmsedge] (sectms) {{\scriptsize\Cref{subsec:cs:tms}}};
      \node[right=0.5 of sectms] (secsms) {{\scriptsize\Cref{subsec:cs:ms}}};
      \node[right=1.75 of secsms] (secopts) {{\scriptsize\Cref{subsec:cs:opts}}};
      \node[right=1.75 of secopts] (secscct) {{\scriptsize\Cref{subsec:cs:scct}}};

      \draw[dashed] ($(sectms)-(0.8,0)$) -- ($(sectms)-(0.8,1.25)$);
      \draw[dashed] ($(sectms)-(-0.9,0)$) -- ($(sectms)-(-0.9,1.25)$);
      \draw[dashed] ($(secsms)-(-1.2,0)$) -- ($(secsms)-(-1.2,1.25)$);
      \draw[dashed] ($(secscct)-(1.2,0)$) -- ($(secscct)-(1.2,1.25)$);
      \draw[dashed] ($(secscct)-(-1.1,0)$) -- ($(secscct)-(-1.1,1.25)$);
    \end{tikzpicture}
  \end{center}
  \caption{Visualisation of the optimizing compilation pipeline that attains a combination of \gls{ms} and \gls{cct}.}\label{fig:pipeline}
\end{figure}
This section defines several secure compilers, each of which robustly preserves a different property of interest (\Cref{fig:pipeline}).
The power of the framework (\Cref{sec:compprop,sec:compcomp}) is demonstrated by composing these compilers for a secure, but optimizing, compilation chain that robustly preserves \gls{ms} and \gls{cct}.
Concretely, the section starts with a base-line definition for the used languages (\Cref{subsec:cs:defs}) and extends that to define languages $\src{L_{TMS}}$ and $\trg{L}$.
For programmers using $\src{L_{TMS}}$, their mental model should be that their code is temporal memory safe and cryptographic constant time.
The former is proven to be enforced by the static semantics of $\src{L_{TMS}}$, where each well-typed program is temporal memory safe.
The latter is completely untyped.
Having languages defined, this section continues with a compiler $\cc{\src{L_{\tmssafe}}}{\trg{L}}$ and proves that $\cc{\src{L_{\tmssafe}}}{\trg{L}}$ robustly preserves \gls{tms} (\Cref{subsec:cs:tms}).
For readability purposes, it defines $\irl{L_{MS}}$ to be exactly the same as $\trg{L}$ and continues to present a compiler $\cc{\trg{L}}{\irl{L_{MS}}}$ that robustly preserves \gls{sms} (\Cref{subsec:cs:ms}).
Now leaveraging the power of the framework that this paper introduces (\Cref{sec:compcomp,sec:compprop}), a compiler $\rtp{\cc{\src{L_{TMS}}}{\irl{L_{MS}}}}{\mssafe}$ is attained.
The next steps in the compilation pipeline are two secure compilers $\cc[_{\gls{dce}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$ and $\cc[_{\gls{cf}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$ that perform \glsfirst{dce} and \glsfirst{cf}, respectively, to model an optimzing pipeline happening somewhere in the compilation chain (\Cref{subsec:cs:opts}).
Finally, this section defines the language $\obj{L_{\scctsafe}}$ that is, in contrast to the other languages, not cryptographic constant time by definition.
Concluding, a compiler $\cc{\irl{L_{MS}}}{\obj{L_{CCT}}}$ that robustly preserves \gls{cct} (\Cref{subsec:cs:scct}) is shown and composed with $\cc{\src{L_{TMS}}}{\irl{L_{MS}}}$ and the optimizing compilers $\cc[_{\gls{dce}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$ and $\cc[_{\gls{cf}}]{\irl{L_{MS}}}{\irl{L_{MS}}}$.
In turn, this yields $\cc{\src{L_{TMS}}}{\obj{L_{CCT}}}$, a compiler that robustly preserves a combination of \gls{ms} and \gls{cct}.

@@

\begin{theorem}[Compiler $\ccb$ is secure with respect to \gls{sms}]\label{thm:ccb:rtp:sms}
  $\rtp{\ccb}{\smssafe}$ % \Coqed
\end{theorem}

@@

\subsection{Optimizing Compilers}\label{subsec:cs:opts}

\begin{center}
  $$
  \begin{array}{rcl}
    \ccdce(\irl{\lifz{true}{\expr[_{1}]}{\expr[_{2}]}}) & = & \ccdce(\irl{\expr[_{1}]}) \\
    \ccdce(\irl{\lifz{false}{\expr[_{1}]}{\expr[_{2}]}}) & = & \ccdce(\irl{\expr[_{2}]}) \\
  \end{array}
  $$
\end{center}

\begin{center}
  $$
  \begin{array}{rcll}
    \cccf(\irl{\llet{x}{n}{\expr}}) & = & \cccf(\irl{\expr\subst{n}{x}}) &\\
    \cccf(\irl{\lbinop{n_{1}}{n_{2}}}) & = & \irl{n_{3}} & \text{where }\irl{n_{3}}\text{ is }\lbinop{\irl{n_{1}}}{\irl{n_{2}}}\\
  \end{array}
  $$
\end{center}

\begin{theorem}[Compiler $\ccdce$ is secure with respect to \gls{ms}]\label{thm:ccdce:rtp:ms}
  $\rtp{\ccdce}{\mssafe}$ % \Coqed
\end{theorem}
\begin{theorem}[Compiler $\cccf$ is secure with respect to \gls{ms}]\label{thm:cccf:rtp:ms}
  $\rtp{\cccf}{\mssafe}$ % \Coqed
\end{theorem}

\begin{theorem}[Compilers $\cccf\circ\ccdce$ and $\cccf\circ\ccdce$ are secure with respect to \gls{ms}]\label{thm:cccfccdce:rtp:ms}
  $\rtp{\cccf\circ\ccdce}{\mssafe}$ and $\rtp{\ccdce\circ\cccf}{\mssafe}$ % \Coqed
\end{theorem}

@@

\begin{center}
  $$
  \begin{array}{rcl}
    \ccscct(\irl{\lfunction{g}{x}{\expr}}) & = & \lfunction[\obj]{\obj{g}}{\obj{x}}{\obj{\lwrdoit{1};}\ccscct(\irl{\expr})}
  \end{array}
  $$
\end{center}

\begin{theorem}[Compiler $\ccscct$ is secure with respect to \gls{scct}]\label{thm:ccscct:rtp:scct}
  $\rtp{\ccscct}{\scctsafe}$ % \Coqed
\end{theorem}

\subsection{Robust Preservation of Intersection of Memory Safety and Strict Cryptographic Constant Time}

Let $\cc{\Ltms}{\Lscct}$ be the compiler that is the composition of $\cca$, $\ccb$, $\cccf$, $\ccdce$, and $\ccscct$, then the following theorem can be shown.

\begin{theorem}[Compiler $\ccscct$ is secure with respect to \gls{scct}]\label{thm:ccscct:rtp:scct}
  $\rtp{\ccscct}{\scctsafe}$ % \Coqed
\end{theorem}
